Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-08-18 surfaces-write-up/surface-write-up.tex
got rid of equation numbers when un-referenced

\maketitle
\todo{Aaron: Remove labels which are not referenced
\\
Peter: I did this for a few of them, but I'm wondering what the
purpose of this is.  It appears the same in the PDF either way
right?

Aaron: Well, the point is to actually remove the references
numbers from equations which aren't referenced, since this
removes clutter. That is, put a star after the environment
name (change begin{equation} to begin{equation*}).
I think it makes sense not to label things
if you don't use those labels, and 
David recommended doing this in his comments.
}
\todo{Aaron: Make sure indexing at 0 is correct in projective space
section, particularly toward the end}
%%%%%%%%%%%%%%%%%%%%%%%%%%%% Introduction %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{proof}
By definition of $H^0(\bp^m,dD)$, functions satisfying conditions 
~\eqref{align:pm-span} lie in $H^0(\mathbb{P}^m,dD)$.
\todo{Aaron: Peter, what do you mean when you say ``all functions of that form lie in...''? How is
this different from saying that functions satisfying that equation lie in $H^0$?
\\
Peter:  Oops, meant to write the converse statement, but accidently
just restated the original statement-it's fixed now.

Aaron: Well, it should really be all such functions are linear
combinations of things in that form. However,
that's what I proved in the next couple setences. I.e.,
the remainder of the proof is showing that this set is indeed
a basis. I removed your added clause.
}
To complete the proof, it
suffices to check functions satisfying conditions ~\eqref{align:pm-basis} form
a basis of $H^0(\bp^m,dD)$. Note that there are $\binom{m+ \deg \lfloor dD \rfloor }{m}$
functions satisfying condition ~\eqref{align:pm-basis}. However we know
$h^0(\bp^m,dD) = \binom{m+ \deg \lfloor dD \rfloor }{m},$ so it suffices to show
that those functions satisfying condition ~\eqref{align:pm-basis} are
independent. This follows from the assumption that $f_0,\ldots f_m$ form a
basis of linear forms, and so all monomials in $f_0,\ldots, f_m$
of degree $\deg \lfloor d \deg D \rfloor $
form a basis of degree $\deg \lfloor d  D \rfloor$ rational functions.
\end{proof}

@@

\begin{lem}
\label{lem:composite-map}
Lex $X$ be an arbitrary scheme and let $D = \sum_{i=0}^{n}\alpha_i D_i\in \bq \otimes_\bz \di X$ where $D_i = V(f_i)$. Suppose we have a
surjection $\phi\colon \bk[w_1,\ldots, w_r] \rightarrow R_D$ given by $w_i
\mapsto p_i(f_0, \ldots, f_n),$ where $p_i$ is a monomial in $f_0,\ldots, f_n$.
Let $\bida_0, \ldots, \bida_n, \bidb_0, \ldots, \bidb_n \in \bz_{\geq 0}.$
Then, define
\begin{align*}
	\Sigma = \langle u^d z_0^{c_0} \cdots z_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=0}^{n} \bida_i c_i = \sum_{i=0}^{n} \bidb_i c_i = 0 \rangle. 
\end{align*}
In this case, we can factor $\phi$ as a composition of $\chi$ and $\psi$ defined by
\[
\begin{tikzcd}
\bk[w_1,\ldots, w_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
w_i \arrow[mapsto]{r} & u^{d_i}z_0^{c_{i0}} \cdots z_n^{c_{in}} \arrow[mapsto]{r} & u^{d_i}f_0^{c_{i0}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$
is at most the maximum of the minimal degree of generation of $\ker \chi$ and
the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
First, we show there is an exact sequence
\[
\begin{tikzcd}
0 \ar {r} & \ker \chi \ar{r}{\rho} & \ker \phi \ar {r}{\tau} & \ker \psi \ar r & 0.
\end{tikzcd}
\]
To see this, note that $\rho$ is injective and $\tau$ is surjective\ because
$\phi = \chi \circ \psi$. Additionally, $\tau \circ \rho = 0$ by definition of
the kernel of $\chi$. Finally, if $a \in\ker \tau$, then by definition $\chi(a)
= 0,$ so $a \in \im \rho$.

@@

From this geometric description of the extremal rays, we can write
the extremal rays algebraically as follows.
For $i \in \Te(D)$, define $e_i \in \bk[\Sigma]$ in degree 
\[
	d_i = \ell_i \gcd(a_i, b_i)
\]
by 
\begin{equation}
	e_i \colon = d_i \epsilon_i %(-\alpha_1 d_i, \ldots, - \alpha_{i-1} d_i, \frac{d_i}{\bida_i + \bidb_i}\sum_{k \ne i}
	%\alpha_k (\bida_k + \bidb_k), - \alpha_{i+1} d_i, \ldots, - \alpha_n d_i)
\end{equation}

\noindent
For $i \in \Tp(D)$ and $j \in \Tm(D)$, with $i < j$, define $e_{i,j}\in \bk[\Sigma]$ in degree 
\[
	d_{i,j} = \ell_{i, j}(\bida_i \bidb_j - \bida_j \bidb_i)
\]
by
\begin{align}
e_{i,j} \colon = &\frac{a_j\sum_{k\ne i,j} d_{i,j}\alpha_k b_k - b_j\sum_{k \ne i,j} d_{i,j} \alpha_k a_k}{a_ib_j - b_i a_j} \epsilon_i \\
&+ \frac{a_i \sum_{k\ne i,j} d_{i,j} \alpha_k b_k - b_i\sum_{k\ne i,j} d_{i,j} \alpha_ka_k}{a_jb_i - b_j a_i} \epsilon_j
%	
%	(-\alpha_1 d_{i,j}, \ldots -\alpha_{i-1} d_{i,j}, \bidb_j \sum_{k \ne i,j} \alpha_k \bida_k - \bida_j \sum_{k \ne i, j}
%	\alpha_k \bidb_k, -\alpha_{i+1} d_{i,j}, \\
%	&\ldots -\alpha_{j-1} d_{i,j}, -\bidb_i \sum_{k \ne i, j} \alpha_k \bida_k + \bida_i \sum_{k \ne i, j}
%	\alpha_k \bidb_k, -\alpha_{j+1}d_{i,j}, \ldots, - \alpha_{n}d_{i,j}).
%	\end{split}
\end{align}

%\noindent
%where
%\[
%	s_1 = \bidb_j \sum_{k \ne i,j} \alpha_k \bida_k - \bida_j \sum_{k \ne i, j}
%	\alpha_k \bidb_k
%\]
%
%\noindent
%and
%\[
%	s_2 = -\bidb_i \sum_{k \ne i, j} \alpha_k \bida_k + \bida_i \sum_{k \ne i, j}
%	\alpha_k \bidb_k.
%\]

@@

\begin{proof}
We first claim that there exist $u^{\deg z_1}\beta_1, \ldots, u^{\deg z_n} \beta_n\in \bk[uz_1, uz_2, uz_3, uz_4]$
such that $\ker \psi$ has relations of the form
\begin{equation}
	u^d(z_i - \beta_i)\prod_{j=1}^n {z_j}^{c_{j}}
\end{equation}

@@

Define
\begin{equation}
	\lambda \colon = \sigma - E = \sum_{j\in \Te} (\lfloor s_j - r_j\rfloor) e_j + \sum_{\substack{j\in \Tp \\ k\in \Tm}} ( \lfloor s_{j,k} - r_{j,k}\rfloor)e_{j,k}\in \Sigma.
 \end{equation}

>>>>

\maketitle
\todo{Aaron: Make sure indexing at 0 is correct in projective space
section, particularly toward the end.
\\
Peter: Think I got all of them.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%% Introduction %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{proof}
By definition of $H^0(\bp^m,dD)$, functions satisfying conditions 
~\eqref{align:pm-span} lie in $H^0(\mathbb{P}^m,dD)$.
To complete the proof, it
suffices to check functions satisfying conditions ~\eqref{align:pm-basis} form
a basis of $H^0(\bp^m,dD)$. Note that there are $\binom{m+ \deg \lfloor dD \rfloor }{m}$
functions satisfying condition ~\eqref{align:pm-basis}. However we know
$h^0(\bp^m,dD) = \binom{m+ \deg \lfloor dD \rfloor }{m},$ so it suffices to show
that those functions satisfying condition ~\eqref{align:pm-basis} are
independent. This follows from the assumption that $f_0,\ldots f_m$ form a
basis of linear forms, and so all monomials in $f_0,\ldots, f_m$
of degree $\deg \lfloor d \deg D \rfloor $
form a basis of degree $\deg \lfloor d  D \rfloor$ rational functions.
\end{proof}

@@

\begin{lem}
\label{lem:composite-map}
Let $X$ be an arbitrary scheme and let $D = \sum_{i=0}^{n}\alpha_i D_i\in \bq \otimes_\bz \di X$ where $D_i = V(f_i)$. Suppose we have a
surjection $\phi\colon \bk[w_1,\ldots, w_r] \rightarrow R_D$ given by $w_i
\mapsto p_i(f_0, \ldots, f_n),$ where $p_i$ is a monomial in $f_0,\ldots, f_n$.
Let $\bida_0, \ldots, \bida_n, \bidb_0, \ldots, \bidb_n \in \bz_{\geq 0}.$
Then, define
\begin{align*}
	\Sigma = \langle u^d z_0^{c_0} \cdots z_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=0}^{n} \bida_i c_i = \sum_{i=0}^{n} \bidb_i c_i = 0 \rangle. 
\end{align*}
In this case, we can factor $\phi$ as a composition of $\chi$ and $\psi$ defined by
\[
\begin{tikzcd}
\bk[w_1,\ldots, w_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
w_i \arrow[mapsto]{r} & u^{d_i}z_0^{c_{i0}} \cdots z_n^{c_{in}} \arrow[mapsto]{r} & u^{d_i}f_0^{c_{i0}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$
is at most the maximum of the minimal degree of generation of $\ker \chi$ and
the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
First, we show there is an exact sequence
\[
\begin{tikzcd}
0 \ar {r} & \ker \chi \ar{r}{\rho} & \ker \phi \ar {r}{\tau} & \ker \psi \ar r & 0.
\end{tikzcd}
\]
To see this, note that $\rho$ is injective and $\tau$ is surjective\ because
$\phi = \chi \circ \psi$. Additionally, $\tau \circ \rho = 0$ by definition of
the kernel of $\chi$. Finally, if $a \in\ker \tau$, then by definition $\chi(a)
= 0,$ so $a \in \im \rho$.

@@

From this geometric description of the extremal rays, we can write
the extremal rays algebraically as follows.
For $i \in \Te(D)$, define $e_i \in \bk[\Sigma]$ in degree 
\[
	d_i = \ell_i \gcd(a_i, b_i)
\]
by 
\[
	e_i \colon = d_i \epsilon_i.
\]

\noindent
For $i \in \Tp(D)$ and $j \in \Tm(D)$, with $i < j$, define $e_{i,j}\in \bk[\Sigma]$ in degree 
\[
	d_{i,j} = \ell_{i, j}(\bida_i \bidb_j - \bida_j \bidb_i)
\]
by
\begin{align*}
e_{i,j} \colon = &\frac{a_j\sum_{k\ne i,j} d_{i,j}\alpha_k b_k - b_j\sum_{k \ne i,j} d_{i,j} \alpha_k a_k}{a_ib_j - b_i a_j} \epsilon_i \\
&+ \frac{a_i \sum_{k\ne i,j} d_{i,j} \alpha_k b_k - b_i\sum_{k\ne i,j} d_{i,j} \alpha_ka_k}{a_jb_i - b_j a_i} \epsilon_j
\end{align*}

@@

\begin{proof}
We first claim that there exist $u^{\deg z_1}\beta_1, \ldots, u^{\deg z_n} \beta_n\in \bk[uz_1, uz_2, uz_3, uz_4]$
such that $\ker \psi$ has relations of the form
\[
	u^d(z_i - \beta_i)\prod_{j=1}^n {z_j}^{c_{j}}
\]

@@

Define
\[
	\lambda \colon = \sigma - E = \sum_{j\in \Te} (\lfloor s_j - r_j\rfloor) e_j + \sum_{\substack{j\in \Tp \\ k\in \Tm}} ( \lfloor s_{j,k} - r_{j,k}\rfloor)e_{j,k}\in \Sigma.
\]
